---
output:
  github_document:
    fig_width: 9
    fig_height: 7
    toc: false
    df_print: kable
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r knitr_init, echo=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  warning=FALSE,
  message=FALSE,
  comment="#>",
  fig.path="man/figures/README-"
);
options(knitr.table.format='markdown')
```

# Jam SummarizedExperiment Stats (jamses)

This package is under active development. These functions
are in frequent use in active Omics analysis projects.
A summary of goals and relevant features are described below.

[Jamses Online Documentation](https://jmw86069.github.io/jamses)

## Goals of jamses

The core goal is to make data analysis and visualization of
`SummarizedExperiment` objects straightforward for
common scenarios.
It also accepts `SingleCellExperiment` and `Seurat` objects.

* **[Make Effective Heatmaps](https://jmw86069.github.io/jamses/articles/how_to_heatmap.html)**
* **Apply Normalization / Adjustment**
* `SEDesign`: **Create Design and Contrast matrices**

   * Use `~ 0 + group` syntax, see below.
   * Integrate Samples, Groups, and Contrasts.
   * Visualize with `plot_sedesign()`.

* `SEStats`: **Analyze Multiple Contrasts**

   * Use limma / limma-voom / limma-DEqMS.

* **Convenient contrast labels**

       Contrast:
       `(Knockout_treated-Knockout_control)-(Wildtype_treated-Wildtype_control)`

       Comp:
       `Knockout-Wildtype:treated-control`
   
* **Integrate with other tools**.

   * `SummarizedExperiment`, `SingleCellExperiment`, `Seurat`,
   `ExpressionSet`, `DESeqDataset`, `DGElist`
   * `venndir::venndir()` - see Github `"jmw86069/venndir"` to create
   directional Venn diagrams.

## Make Effective Heatmaps

[How to Make a Heatmap](https://jmw86069.github.io/jamses/articles/how_to_heatmap.html)

Jamses `heatmap_se()` reinforces heatmap principles, and
applies some opinions.

* Data are centered.
* Always use divergent colors for centered data.
* Data are **not scaled**.
* Row and Column annotations are "easy".


<details><summary>
## Approach for Statistical Contrasts
</summary>

[The Limma User's Guide](https://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/)
(LUG) is an amazing resource which describes numerous approaches for
one-way and two-way contrasts **which are mathematically equivalent**.
For a more thorough discussion please review these approaches to confirm
that `~0 + x` is mathematically identical to `~ x`, and only differs
in how estimates are reported.

* Jamses uses the `~0 + x` strategy.
* Each experiment group is defined using independent replicates.<br>
* This approach *does not imply* that there is "no intercept" during
the model fit, see LUG for details.
* One-way contrasts compare only one factor per contrast.

   * Valid one-way contrast:
   ```
   (A_treated - A_control) # valid one-way contrast
   ```

   * Invalid one-way contrast:
   ```
   (A_treated - B_control) # not a valid one-way contrast
   ```

* Two-way contrasts in Jamses compare the fold change of two
***compatible*** one-way fold changes.

   * Two compatible one-way contrasts:
   ```
   (A_treated - A_control)  # one-way contrast
                            # "treated-control" for A
   (B_treated - B_control)  # compatible one-way contrast
                            # "treated-control" for B
   ```

   * Corresponding two-way contrast:
   ```
   (B_treated - B_knockout) # incompatible one-way contrast
   ```

</details>

## SEDesign: Design and Contrasts

* `groups_to_sedesign()` takes by default a `data.frame` where each column
represents an experiment factor, and creates the following:
* Output is `SEDesign` as an S4 object with slot names:
   * `design(sedesign)` - the design matrix.
   * `contrasts(sedesign)` - the contrasts matrix.
   * `samples(sedesign)` - vector of samples.


### Example SEDesign object

The example below uses a `character` vector of group names per sample,
with two factors separated by underscore `"_"`.
The same data can be provided as a `data.frame` with two columns.

```{r sedesign, collapse=TRUE}
library(jamses)

igroups <- jamba::nameVector(paste(rep(c("WT", "KO"), each=6),
   rep(c("Control", "Treated"), each=3),
   sep="_"),
   suffix="_rep");
igroups <- factor(igroups, levels=unique(igroups));

jamba::kable_coloring(color_cells=FALSE,
   caption="Sample to group association",
   data.frame(groups=igroups))
```

The resulting design and contrasts matrices are shown below:

```{r sedesign_output, collapse=TRUE}
sedesign <- groups_to_sedesign(igroups);
jamba::kable_coloring(
   colorSub=c(`-1`="dodgerblue", `1`="firebrick"),
   caption="Design matrix output from design(sedesign).",
   data.frame(check.names=FALSE, design(sedesign)));

jamba::kable_coloring(
   colorSub=c(`-1`="dodgerblue", `1`="firebrick"),
   caption="Contrast matrix output from contrasts(sedesign).",
   data.frame(check.names=FALSE, contrasts(sedesign)));

```

For convenience, SEDesign can be visualized using `plot_sedesign()`:

```{r plot_sedesign, collapse=TRUE, alt="Plot of the 'SEDesign' object, using squares to indicate groups, and block arrows to indicate contrasts."}
# plot the design and contrasts
plot_sedesign(sedesign);
title(main="plot_sedesign(sedesign):")
```

* One-way contrasts are shown with a wide block arrow.
* Two-way contrasts are shown by connecting two block arrows
with a "squiggly curved line".

   * It connects the end of one contrast   
   to the beginning of the next contrast.
   * The order indicates that the first contrast
   is subtracted by the second contrast.

