---
title: "How to Make a Heatmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Make a Heatmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev="ragg_png",
  fig.height=9,
  fig.width=8,
  fig.retina=2,
  dpi=96
)
```

```{r setup}
library(jamses)
```


The `jamses` package provides `heatmap_se()` which is intended
to "make heatmaps easy" when used with a `SummarizedExperiment`
object.

You can use a `numeric` matrix, but it won't have many of the
fancy extra features.

## About SummarizedExperiment Objects

(See [SummarizedExperiment vignette](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html)
for more details on the object design.)

A `SummarizedExperiment` is just a fancy matrix, with these additions:

* It can store more than one `numeric` matrix using `assays()`.
* Column annotations are stored with `SummarizedExperiment::colData()`.
* Row annotations are stored with `SummarizedExperiment::rowData()`.

It helps associate rich detail to the columns (biological samples)
and rows (genes or platform measurements).

Column and Row annotations are used in `heatmap_se()`

* to add annotations alongside the heatmap
* to split the heatmap by useful criteria
* to center the data in meaningful ways

Several data types can be converted to `SummarizedExperiment`:
`ExpressionSet`, `DGEList`, `DESeqDataset`, `matrix`.


## Design Philosophy

The `heatmap_se()` is "opinionated".
Default behavior should make a heatmap very close to ideal.

Most opinions have exceptions.  
(There are a lot of exceptions.)

We start with `ComplexHeatmap::Heatmap()`. It is amazing.
(And complex. Ha!)

`heatmap_se()` tries to make it easy.

### Default Assumptions

* Data are assumed to be log2-transformed. Usually `log2(1 + x)`, ymmv.

* **Data are not row-scaled.** *(Gasp)*

   * We believe magnitude of change is important
   in Omics data.
   It is helpful for critical review of the data,
   which is the main purpose of a heatmap.
   * Row-scaling shows patterns scaled to the
   variability on each row.
   It is particularly effective when the range of
   response is different for each row of data.
   Most Omics data platforms do not have this problem,
   and in fact the magnitude of change is more associated
   with the biology than to the technology.
   * Row-scaling has the nice benefit that it does not
   require log2-transformed data. It "naturally" transforms
   data into "variance" (z-score) units.
   These units may not be biologically meaningful.
   * It attempts to allow large and small magnitude effects
   to have similar influence, which is an important goal.
   We aim to achieve this goal without row-scaling.

* Data are row-centered.

   * The average signal of each row is subtracted from
   the row, revealing "difference from average".
   * When using log2-transformed data, the values are
   "log2 differences".
   * These values are equivalent to log2 fold
   changes as shown in a volcano plot, and equivalent to
   differences as shown in MA-plots.
   * Why not display a heatmap using values that are
   consistent with those used for statistical tests,
   used in other figures?

* **Always use a divergent color scale with centered data.**

   * The color scale is blue-white-red, white at zero.
   * <span style='color:red'>**Red is Up**</span> because this is a
   <span style='color:red'>**HEAT map**</span>.
   * <span style='color:blue'>**Blue is Down**</span> because it is
   <span style='color:blue'>**cold**</span>.
   * When we see a heatmap with <span style='color:blue'>**Blue**</span>
   as the high color, we refer to it as a
   <span style='color:blue'>*COLD map*</span>.  
   You can make a cold map, please do not call it a heat map.

* The color scale uses a fixed range.

   * The default range: **-8 to +8 fold**.
   * The default heatmap has consistent color-to-magnitude
   relationship.
   * We use this consistent color profile because it fits
   biologically relevant changes for **most** Omics data.
   * Extremely high values do not change this range,
   otherwise it would compress the colors of everything else.
   * When the data have small changes, the heatmap will show
   them as small changes.
   * The range may be expanded when needed.

* Data centering is flexible:

   * By default, all samples are the controls.
   * Centering can be "versus controls" using reference
   samples.
   * Centering can be performed within independent sub-groups.
   We call them `'centerby groups'`.
   * Examples of `'centerby groups'`: 
   design factor(s),
   sample type,
   tissue type,
   cell line,
   processing batch,
   pairing factor,
   * Each `'centerby group'` can have its own controls.

* Annotations are "easy":

   * `top_colnames` adds `colData()` annotations at the top.
   * `rowData_colnames` shows `rowData()` annotations on the left.

* Statistical hits are "easy":

   * `sestats` adds statistical hits to the left.
   * Hits are colored by direction.
   * The heatmap is subset to show only these hits.

* Rows and columns can be grouped:

   * `row_split` defines row groups
   * `column_split` defines column groups
   * use an `integer` number of groups, or
   annotation `colnames()`.


# Heatmap Walkthrough

## Test Data

We generate test data for this demonstration.

```{r hm_test_data}
se <- make_se_test(nrow=1000, ngroups=4, nreps=8)

# optionally define factor levels to force the order of labels
SummarizedExperiment::rowData(se)$Class <- factor(
   sample(head(LETTERS, 5), size=nrow(se), replace=TRUE))
```

## Basic Heatmap

```{r hm_basic}
hm <- heatmap_se(se, rowData_colnames="Class")

# draw by printing hm, or call draw() to add useful options
ComplexHeatmap::draw(hm,
   column_title=attr(hm, "hm_title"),
   merge_legends=TRUE)
```

## Custom Colors

```{r hm_colors}
sample_color_list <- list(
   group=c(
      groupA="gold",
      groupB="darkorange2",
      groupC="firebrick3",
      groupD="darkorchid4"
   ),
   Class=colorjam::group2colors(
      unique(SummarizedExperiment::rowData(se)$Class)))

heatmap_se(se,
   rowData_colnames="Class",
   show_left_annotation_name="top",
   left_annotation_name_rot=150,
   sample_color_list=sample_color_list)
```

## Heatmap Legend Options

`ComplexHeatmap::ht_opt()` allows you to customize some stylistic
defaults. One common option is: `merge_legends=TRUE` which
combines the annotation and heatmap color legends so they
appear together in one column.

```{r hm_merge_legends}
ComplexHeatmap::ht_opt(merge_legends=TRUE)

heatmap_se(se,
   rowData_colnames="Class",
   show_left_annotation_name="top",
   left_annotation_name_rot=150,
   sample_color_list=sample_color_list)

```

## Split Rows by Annotation

```{r hm_row_split}
heatmap_se(se,
   rowData_colnames="Class",
   row_split="Class",
   sample_color_list=sample_color_list)
```

## Split Rows and Columns Together

```{r hm_row_column_split}
hm2 <- heatmap_se(se,
   column_split=c("group"),
   column_title_rot=90,
   row_split=c("Class"),
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
hm2drawn <- ComplexHeatmap::draw(hm2, merge_legends=TRUE)
```

```{r hm_row_order}
# as an example, extract the row order
# technically you should use hm2drawn, but usually hm2 is enough
hro <- jamba::heatmap_row_order(hm2drawn);
jamba::sdim(hro)
lapply(hro, head, 7)
# (the names will differ from values when `row_labels` are customized)
```

## Center Using a Control Group

This heatmap calculates the mean for the control
samples `controlSamples`, then uses that value
for data centering.

```{r hm_control_group}
# center by WildType samples
# - controlSamples
# - control_label
hm2 <- heatmap_se(se,
   controlSamples=rownames(subset(
      SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs groupA",
   column_split=c("group"),
   column_title_rot=90,
   row_split=c("Class"),
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
hm2drawn <- ComplexHeatmap::draw(hm2,
   column_title=attr(hm2, "hm_title"),
   merge_legends=TRUE)
```

## Add Callout Labels for a Subset of Rows

You can label a subset of genes of interest,
useful especially when there are too many rows
to label everything.

```{r hm_mark_rows}
# add "callout" labels for a subset of rows
mark_rows <- c(sample(jamba::heatmap_row_order(hm2drawn)[[1]], size=5),
   sample(jamba::heatmap_row_order(hm2drawn)[[1]], size=3));

# turn off ComplexHeatmap warning when using RStudio
ComplexHeatmap::ht_opt(message=FALSE)

hm3 <- heatmap_se(se,
   mark_rows=mark_rows,
   controlSamples=rownames(
      subset(SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs groupA",
   column_split=c("group"),
   column_title_rot=90,
   row_split=c("Class"),
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
ComplexHeatmap::draw(hm3,
   column_title=attr(hm3, "hm_title"),
   merge_legends=TRUE)
```

## Display Statistical Hits

The argument `sestats` is used to display statistical
hits on the left of the heatmap.
It can be one of a few data types:

* `list` named by contrast, containing `numeric` vectors

   * Each `numeric` vector is named by `gene`, which
   should match `rownames()` of the data.
   * Any non-zero value is considered a "hit", and usually
   the vector contains the sign `1` or `-1`.

* `numeric` matrix where `rownames()` match the data,
and `colnames()` are used for each comparison.
* `SEStats` object produced by `se_contrast_stats()`.

```{r hm_sestats_list}
# sestats can accept list, incidence matrix, hit_array, or sestats
# this example defines random set of hits
sestats_list <- list(
   contrast1=setNames(sample(c(1, -1), replace=TRUE, size=50),
      sample(rownames(se), size=50)),
   contrast2=setNames(sample(c(1, -1), replace=TRUE, size=50),
      sample(rownames(se), size=50)))
hm4 <- heatmap_se(se,
   controlSamples=rownames(
      subset(SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs groupA",
   sestats=sestats_list,
   column_split=c("group"),
   row_split=c("Class"),
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
ComplexHeatmap::draw(hm4,
   column_title=attr(hm4, "hm_title"),
   merge_legends=TRUE)
```

## Run Limma Statistics with `se_contrast_stats()`

The `se_contrast_stats()` function is a wrapper to `limma`,
and its supporting functions `lmFit()`, `fitContrasts()`,
`topTable()`, and even `voom()` when appropriate.
It also generates stats tables for each contrast in
one step, then annotates statistical hits that meet
the cutoff criteria.

The output is `SEStats` object, which is a convenient way
to store results from multiple statistical contrasts.
It can also perform contrasts with multiple `assay_names`.

```{r hm_se_contrast_stats}
# it doesn't take much effort to run stats really quick
sedesign <- groups_to_sedesign(SummarizedExperiment::colData(se)[, "group", drop=FALSE])
contrast_names(sedesign) <- jamba::vigrep("-groupA", contrast_names(sedesign))
sestats <- se_contrast_stats(se=se,
   fold_cutoff=4,
   sedesign=sedesign, assay_name="counts")
hm4s <- heatmap_se(se,
   controlSamples=rownames(
      subset(SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs groupA",
   sestats=sestats,
   column_split=c("group"),
   row_split=6,
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
ComplexHeatmap::draw(hm4s,
   column_title=attr(hm4s, "hm_title"),
   merge_legends=TRUE)
```

## Drill Down into a Row Cluster

When clustering rows, it is sometimes useful
to look at a particular row subcluster to see
more detail, perhaps just to read the row labels.
The argument `row_subcluster` does the work of
determining the row order, and generating a new
heatmap with the same row order, labeled using
the same cluster names.

```{r hm_row_subcluster}
# for fun, "drill down" into cluster 5
hm4s_4 <- heatmap_se(se,
   controlSamples=rownames(
      subset(SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs groupA",
   sestats=sestats,
   column_split=c("group"),
   row_split=6,
   row_subcluster=4,
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
ComplexHeatmap::draw(hm4s_4,
   column_title=attr(hm4s_4, "hm_title"),
   merge_legends=TRUE)
```

## Use an Incidence Matrix for `sestats`

```{r hm_sestats_incidence, eval=jamba::check_pkg_installed("venndir")}
# sestats can be provided as an incidence matrix
# convert sestats to list
sestats_hitlist <- hit_array_to_list(sestats)
# convert sestats hitlist to incidence matrix
# - for fun, use only the first two contrasts
sestats_hitim <- venndir::list2im_value(sestats_hitlist[1:2])
print(head(sestats_hitim));

# convert sestats_list to signed incidence matrix
sestats_im <- venndir::list2im_value(sestats_list)
print(head(sestats_im, 10));
# if the list has items (no direction) use venndir::list2im_opt()

hm5 <- heatmap_se(se,
   controlSamples=rownames(
      subset(SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs groupA",
   sestats=sestats_hitim,
   column_split=c("group"),
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
ComplexHeatmap::draw(hm5,
   column_title=attr(hm5, "hm_title"),
   merge_legends=TRUE)
```

## Customize Column Label Fonts

This option is rarely used, but sometimes helpful.
The column labels can be customized to use a custom font,
bold fontface, or even custom colors.

```{r hm_column_fonts}
# customize column label fonts using column_names_gp
column_bold <- ifelse(
   SummarizedExperiment::colData(se)$group %in% "groupA",
   2, 1);
hm6 <- heatmap_se(se,
   controlSamples=rownames(
      subset(SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs WildType",
   column_names_gp=grid::gpar(col=sample_color_list$group[
      as.character(SummarizedExperiment::colData(se)$group)],
      font=column_bold),
   column_split=c("group"),
   row_split=c("Class"),
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
ComplexHeatmap::draw(hm6,
   column_title=attr(hm6, "hm_title"),
   merge_legends=TRUE)
```

## Create a Correlation Heatmap

Any heatmap can be displayed as a "correlation heatmap".
The correlation matrix uses the same data that would
have been displayed as the centered heatmap, except the
centered data are used in the correlation `cor()` method.

Using centered data for correlation improves the sensitivity
especially with biological data. It determines correlation
using "difference from average" which tells whether any
changes from average are consistent within sample groups,
or across sample groups.

A good rule of thumb, if you see
a "shadow diagonal line" in this plot, it can be one indication
of a sample pairing effect. To test a sample pairing issue,
you may want to use `centerby_colnames` with the pairing factor
column name.

```{r hm_correlation}
# correlation=TRUE, any heatmap becomes a sample correlation heatmap
hm6corr <- heatmap_se(se,
   correlation=TRUE,
   apply_hm_column_title=TRUE,
   controlSamples=rownames(
      subset(SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs groupA",
   column_names_gp=grid::gpar(col=sample_color_list$group[
      as.character(SummarizedExperiment::colData(se)$group)],
      font=rep(c(1, 2, 1), c(3, 5, 24))),
   column_split=c("Group"),
   sample_color_list=sample_color_list)
ComplexHeatmap::draw(hm6corr,
   merge_legends=TRUE)
```

## Final Heatmap with Grouped Column Labels

The final heatmap combines several conveniences:

1. `apply_hm_column_title=TRUE` applies a heatmap title as `column_title`.
2. `top_colnames` hides the  annotation.
3. `heatmap_column_group_labels()` adds grouped labels
above the heatmap.
4. `hm_title_buffer` adjusts the whitespace for the title and grouped lines.


```{r hm_final}
SummarizedExperiment::colData(se)$Genotype <- rep(c("WT", "KO"), each=16);
SummarizedExperiment::colData(se)$Treatment <- rep(c("Control", "Dex"), each=8);
hm7 <- heatmap_se(se,
   apply_hm_column_title=TRUE,
   hm_title_buffer=3,
   controlSamples=rownames(
      subset(SummarizedExperiment::colData(se), group %in% "groupA")),
   control_label="vs groupA",
   sestats=sestats_list,
   top_colnames=FALSE,
   column_split=c("group"),
   row_split=c("Class"),
   rowData_colnames=c("Class"),
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)
hm7_drawn <- ComplexHeatmap::draw(hm7,
   merge_legends=TRUE)

# now add fancy labels
heatmap_column_group_labels(
   hm_group_list=c("Treatment", "Genotype"),
   se=se,
   hm_drawn=hm7_drawn)
# Note: this step does not work consistently inside RStudio plot pane,
# in that case call dev.new() then run the step above to create hm7_drawn,
# then repeat the step below
#
# adjust the height of labels with argument y_offset_lines
# with positive values (upward), or negative values (downward).
```


