---
title: "How to Make a Heatmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Make a Heatmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
   collapse = TRUE,
   comment = "#>",
   dev = "ragg_png",
   fig.height = 8,
   fig.width = 8,
   fig.asp = 1,
   fig.retina = 2,
   dpi = 96
)
```

Load R packages:

```{r setup, warn=FALSE}
library(jamses)
suppressPackageStartupMessages(
library(SummarizedExperiment)
)
suppressPackageStartupMessages(
library(ComplexHeatmap)
)
```

Make the heatmap:

```{r make_hm, eval=FALSE}
heatmap_se(se)
```

The defaults are very good.  
The custom options are great.

# Why SummarizedExperiment?

<details><summary>**`SummarizedExperiment` is an enhanced data matrix**</summary>

* It stores multiple data matrices.  
Some examples:
   * Raw
   * normalized
   * batch-adjusted
   * VST-normalized
   * "counts" (or pseudocounts)
   * "abundance" (or TPM, FPKM)

* Columns have annotations in `colData()`.
* Rows have annotations in `rowData()`.
* Annotations are used in `heatmap_se()`:
   * display annotations beside the heatmap
   * split the heatmap by groups
   * center the data in meaningful ways

* You can create a `SummarizedExperiment` from any of:
   * `numeric` matrix
   * `ExpressionSet`
   * `DGEList`
   * `DESeqDataset`

</details>

See the [SummarizedExperiment vignette](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html)
for more detail.

# Heatmap Principles

<details><summary>**The `heatmap_se()` is opinionated.**</summary>

* The default should be very close to ideal.  
* *There are many custom options.*

</details>
<details><summary>**We build upon `ComplexHeatmap::Heatmap()`.**</summary>

* It is amazing. And complex. We know this.
* You can combine heatmaps:
   * Add them `+` to display side-by-side.
   * Use `%v%` to display top-to-bottom.

</details>
<details><summary>**Data are assumed to be log2-transformed.**</summary>

* We recommend `log2(1 + x)`, *ymmv*.
* log2-transformed values are consistent with many  
'Omics tools which report log2 fold change.

</details>
<details><summary>**Data are not scaled.** *(Gasp)*</summary>

* We believe magnitude of change is important
in Omics data.  
It is helpful for critical review of the data,  
which is the main purpose of a heatmap.  
(A heatmap is not itself a decision-making tool,  
however it very much helps assess the assumptions  
and results obtained from other tools.)
* Row-scaling adjusts the signal based upon the  
variability on each row.  
It is particularly effective when the range of  
response is different for each row of data,  
**and** where the range of response is more closely  
associated with technical limitations than with  
biological response.  
In our experience, most 'Omics platforms are not
dominated by this limitation.
* Row-scaling has the nice benefit that it does not  
require log2-transformed data.  
It "naturally" transforms data into z-score units.  
These units may or may not be biologically meaningful.
* The z-score units are particularly useful when you  
only want to show presence/absence, or high/low  
relative to the range of signal on each row.
* We aim to achieve many of the same goals without  
scaling, because we find the actual magnitudes useful.

</details>
<details><summary>**Data are row-*centered*.**</summary>

* The average signal of each row is subtracted from  
that row, resulting "difference from average".
* We assume log2-transformed data, therefore values are  
called "log2 differences".
* These log2 differences are commonly shown in volcano plots,  
and in MA-plots.
* The "log2 differences" can be converted to fold changes,  
so we use fold change to label the color key.
* We recommend using the heatmap to display data as  
similar as possible to the underlying statistical tests,  
consistent with other supporting figures.

</details>
<details><summary>**Always use a divergent color scale with centered data.**</summary>

* *(The same rule applies to z-scores, and correlations.)*
* The color scale uses Brewer colors `'RdBu'` in reverse:  
<span style='color:dodgerblue'>blue</span>-white-<span style='color:firebrick'>red</span>  
and white is always at zero.
* <span style='color:firebrick'>**Red is Up**</span> because this is a
<span style='color:firebrick'>**HEAT map**</span>.
* <span style='color:dodgerblue'>**Blue is Down**</span> because it is
<span style='color:dodgerblue'>**cold**</span>.
* If 
<span style='color:dodgerblue'>**Blue were up**</span>,
you should call it a <span style='color:dodgerblue'>***COLD map***</span>.  
A cold map can be useful, but it is not a heat map.

</details>
<details><summary>**The color scale uses a fixed range.**</summary>

* Default range:  
**-8 to +8 fold**.
* It defines a consistent color-to-magnitude relationship.  
Two-fold should look like a two-fold change.
* This range represents biologically relevant changes for  
**most** Omics data. (It can be adjusted.)
* Extremely high values do not adjust the range,  
otherwise it would compress the colors of everything else.
* When the data have small changes, the heatmap will show  
them as small changes.
* The range may be adjusted when needed.

</details>
<details><summary>**Data centering is flexible, and encouraged:**</summary>

* By default, centering uses all samples.  
"Global centering"
* Centering can be "versus controls".
* Centering can be within sub-groups, `'centerby groups'`.  
Examples of `'centerby groups'`:
   * design factor(s),
   * sample type,
   * tissue type,
   * cell line,
   * processing batch,
   * pairing factor,

* Each `'centerby group'` can have its own controls.
* For example, you may center within each cell line,  
then center each sample to its respective control group.

</details>
<details><summary>**Annotations are "easy":**</summary>

* `top_colnames` adds `colData()` annotations at the top.
* `rowData_colnames` shows `rowData()` annotations on the left.
* It will assign categorical colors, or scalar colors,  
or divergent colors.
* You can assign colors every to column individually.

</details>
<details><summary>**Statistical hits are "easy":**</summary>

* Provide `sestats` to add statistical hits to the left side.
* Hits are colored by direction (<span style="color:firebrick">Up</span>/<span style="color:dodgerblue3">Down</span>).
* The heatmap will subset to show only these hits.

</details>
<details><summary>**Rows and columns can be grouped:**</summary>

* `row_split` defines row groups
* `column_split` defines column groups
* use an `integer` number of groups, or
annotation `colnames()`.

</details>

# Heatmap Walkthrough

Each step in the walkthrough adds some detail.
* Sometimes we demonstrate other custom options as well,  
just to give a flavor for the variety of styles available.

## Test Data

We generate test data for this demonstration.

```{r hm_test_data}
se <- make_se_test(nrow=1000, ngroups=4, nreps=8)

# optionally define factor levels to force the order of labels
rowData(se)$Class <- factor(
   sample(head(LETTERS, 5), size=nrow(se), replace=TRUE))
```

## Basic Heatmap

Provide `se` (SummarizedExperiment data) and it will create
the default heatmap.
* `top_colnames` are auto-detected by default.

```{r hm_basic}
hm <- heatmap_se(se)

# draw by printing hm, or call draw() to add useful options
draw(hm)
```

## Custom Annotation Colors

* Use `sample_color_list` to provide custom colors.  
   * Provide a `list` named by the annotation column.  
   * Or use `colorjam::group2colors()` to define categorical colors.

* `rowData_colnames` added row annotations to the left.
* `left_annotation_name_rot=60` rotated the annotation label.

```{r hm_colors}
sample_color_list <- list(
   group=c(
      groupA="gold",
      groupB="darkorange2",
      groupC="firebrick3",
      groupD="darkorchid4"
   ),
   Class=colorjam::group2colors(
      unique(rowData(se)$Class)))

heatmap_se(se,
   rowData_colnames="Class",
   left_annotation_name_rot=60,
   sample_color_list=sample_color_list)
```

## Heatmap Legend Options

* `ComplexHeatmap::ht_opt()` allows some stylistic options  
to be default whenever you draw a heatmap.  
Some common options:
   * `merge_legends=TRUE` merges the color legends to one column.
   * `ROW_ANNO_PADDING`, `COLUMN_ANNO_PADDING` defines a gap  
   between heatmap and row/column annotations, respectively.

* Also: `show_left_annotation_name="top"` places the left  
annotation label at the top instead of the bottom.
* Also: `heatmap_se()` will draw the heatmap by default,  
useful when you don't need to store the output.  
(I usually end up needing the heatmap.)

```{r hm_merge_legends}
ComplexHeatmap::ht_opt(merge_legends=TRUE)

heatmap_se(se,
   rowData_colnames="Class",
   show_left_annotation_name="top",
   left_annotation_name_rot=150,
   sample_color_list=sample_color_list)

```

# Row and Column Groups

## Split Rows by Annotation

* `row_split="Class"` will group rows by "Class".

```{r hm_row_split}
hm <- heatmap_se(se,
   rowData_colnames="Class",
   row_split="Class",
   sample_color_list=sample_color_list)

draw(hm)
```

## Split Rows and Columns

* `column_split="group"` splits the columns.
* `row_split="Class"` splits the rows.
* You can provide multiple columns if needed.

```{r hm_row_column_split}
hm <- heatmap_se(se,
   column_split="group",
   row_split="Class",
   rowData_colnames="Class",
   sample_color_list=sample_color_list)

hm2 <- draw(hm)
```

## Get the Row Group Order

* Now seems like a good time to get the row order:
   * `jamba::heatmap_row_order(hm)`
   * It returns a `list` separated by row group.

* It works best with the `hm2` drawn heatmap, since  
some clustering methods use a random element.

```{r hm_row_order}
hro <- jamba::heatmap_row_order(hm2);
lengths(hro)
```

See the first five entries per row group:
```{r hm_row_order_first5}
lapply(hro, head, 5)
```


```{r hm_se_contrast_stats_prep_hidden, echo=FALSE}
colData(se)$Genotype <- rep(
   factor(c("WT", "KO"), levels=c("WT", "KO")),
   each=16);
colData(se)$Treatment <- rep(
   c("Control", "Dex"), each=8);

# New colors
sample_color_list$Genotype <- c(
   WT="palegoldenrod",
   KO="firebrick")
sample_color_list$Treatment <- c(
   Control="skyblue",
   Dex="navy")

# SEDesign
sedesign <- groups_to_sedesign(se,
   max_depth=1,
   group_colnames=c("Genotype", "Treatment"))

# run stats
sestats <- se_contrast_stats(
   se=se,
   fold_cutoff=4,
   sedesign=sedesign,
   assay_name="counts")

ComplexHeatmap::ht_opt(message=FALSE)
```

## Add Grouped Column Labels

The example illustrates a two-factor design:
1. Genotype: wildtype (WT) and knockout (KO)
2. Treatment: Control and Dexamethasone (Dex)

* Column split: `column_split=c("Genotype", "Treatment")`
* `controlSamples` uses "WT" and "Control"

* Several features are enabled:
   * `apply_hm_column_title=TRUE` displays the heatmap title.
   * `top_colnames=FALSE` hides the top annotation.
   * `heatmap_column_group_labels()` adds grouped labels.
   * `hm_title_buffer` adds whitespace for the grouped lines.
* We add two design factors: `Genotype`, `Treatment`

```{r hm_final}
colData(se)$Genotype <- rep(
   factor(c("WT", "KO"), levels=c("WT", "KO")),
   each=16);
colData(se)$Treatment <- rep(
   c("Control", "Dex"),
   each=8);

# center by groupA samples
use_controlSamples <- rownames(
   subset(colData(se),
      Genotype %in% "WT" &
      Treatment %in% "Control"))

hm9 <- heatmap_se(se,
   apply_hm_column_title=TRUE,
   hm_title_buffer=4,
   controlSamples=use_controlSamples,
   control_label="vs WT_Control",
   sestats=sestats,
   top_colnames=c("Genotype", "Treatment"),
   column_split=c("Genotype", "Treatment"),
   row_split=6,
   column_gap=grid::unit(c(1, 2, 1), "mm"),
   sample_color_list=sample_color_list)
hm9_drawn <- ComplexHeatmap::draw(hm9)

# now add fancy labels
heatmap_column_group_labels(
   hm_group_list=c("Treatment", "Genotype"),
   font_cex=1.3,
   se=se,
   hm_drawn=hm9_drawn)

```

* Use `y_offset_lines` to adjust the y position of labels.


## Change Column Group Order

The example below shows the same heatmap  
grouped by:
1. Treatment
2. Genotype

```{r hm_final_part_deux}
hm10 <- heatmap_se(se,
   apply_hm_column_title=TRUE,
   hm_title_buffer=4,
   controlSamples=use_controlSamples,
   control_label="vs WT_Control",
   sestats=sestats,
   top_colnames=c("Treatment", "Genotype"),
   column_split=c("Treatment", "Genotype"),
   row_split=6,
   sample_color_list=sample_color_list)
hm10_drawn <- ComplexHeatmap::draw(hm10,
   merge_legends=TRUE)

# now add fancy labels
heatmap_column_group_labels(
   hm_group_list=c("Genotype", "Treatment"),
   font_cex=1.2,
   se=se,
   hm_drawn=hm10_drawn)
```




# Data Centering

* Default "data centering" subtracts the row mean value,  
resulting in "difference from global mean".  
We refer to this approach as "global centering".
* Centering typically uses the mean value (the average),  
however it can use median with `useMedian=TRUE`.  
The median is effective in reducing the visual effects  
of outlier points.
* "**Versus control**": Centering can be focused on a  
control group, instead of global centering. This approach  
is very effective at visualizing the data as it is seen  
by statistical contrasts. In fact, it is encouraged.
* "**Within Groups**": Data can be centered within  
sub-groups, called '**center-by groups**'. This approach  
is quite useful and versatile.
   1. Consider an experiment with two different cell lines,  
   comparing treatment versus control.  
   The cell lines have substantial differences from each other,  
   however the treatment effects may be similar.  
   We would "center-by Cell Line", then "versus Control".
   2. Consider an experiment with cells derived from clinical  
   patients, each patient sample is treated multiple ways.  
   Again, "center-by Patient", then either "global center" or  
   "versus Control".

Working examples are shown below.

## Center Using a Control Group

* `controlSamples` defines a subset of samples  
to be the reference for data centering.
   * Here we use samples in `"groupA"`.
   * `control_label="vs GroupA"` describes the control

* `column_title_rot=90` also rotates the column group  
title 90 degrees.
* We define a new attribute `'hm_title'` as a  
convenient heatmap title.
   * `column_title` adds the title when we `draw()`  
   the heatmap.

```{r hm_control_group}
# center by groupA samples
use_controlSamples <- rownames(
      subset(colData(se), group %in% "groupA"))

# - control_label
hm3 <- heatmap_se(se,
   controlSamples=use_controlSamples,
   control_label="vs groupA",
   column_split="group",
   column_title_rot=90,
   row_split="Class",
   rowData_colnames="Class",
   cluster_row_slices=FALSE,
   sample_color_list=sample_color_list)

hm4 <- ComplexHeatmap::draw(hm3,
   column_title=attr(hm3, "hm_title"))
```


## Center By Genotype

* `centerby_colnames="Genotype"` centers within WT or KO
* `use_controlSamples_Treat` uses 'Control' samples  
in both WT and KO.
   * It shows "Dex - Control" in each Genotype.

* `column_gap` adds a custom gap between column groups.

```{r hm_centerby_genotype}
use_controlSamples_Treat <- rownames(
      subset(colData(se), Treatment %in% "Control"))

hm11 <- heatmap_se(se,
   apply_hm_column_title=TRUE,
   hm_title_buffer=3,
   controlSamples=use_controlSamples_Treat,
   control_label="vs Control",
   sestats=sestats,
   top_colnames=FALSE,
   centerby_colnames="Genotype",
   column_split=c("Genotype", "Treatment"),
   row_split=6,
   column_gap=grid::unit(c(2, 4, 2), "mm"),
   sample_color_list=sample_color_list)
hm11_drawn <- ComplexHeatmap::draw(hm11,
   merge_legends=TRUE)

# now add fancy labels
heatmap_column_group_labels(
   hm_group_list=c("Treatment", "Genotype"),
   font_cex=1.2,
   se=se,
   hm_drawn=hm11_drawn)
```


## Center by Batch

* `simulate_se_test()`
   * Simulates test data that contains a blocking effect.  
   It also represents pairing factors (repeated measures),
   or batch effect.
   * When simulating four groups (default),  
   it simulates two-factor additive changes.

* `centerby_colnames="batch"`
> **Important Note:** This technique is entirely visual,  
although it helps confirm potential structure within the data.  
The math very closely mimics the adjustment performed in most  
statistical tools, however we recommend using the proper  
statistical adjustments, specifically `block` in limma, or  
using an additional factor in DESeq2 or edgeR. It is generally  
not recommended to perform statistical testing with  
batch-adjusted data.

### Without Batch-Centering

```{r hm_centerby_batch_1}
seb <- simulate_se_test(ngroups=4, nreps=8, multiplier=2)
colData(seb)[, 2:3] <- colData(se)[, 3:4]
colnames(colData(seb))[2:3] <- colnames(colData(se))[3:4]

hm11 <- heatmap_se(
   seb,
   column_title="Global Centering",
   top_colnames=c("Genotype", "Treatment", "batch"),
   column_split=c("Genotype", "Treatment"),
   row_split=6,
   column_gap=grid::unit(c(2, 4, 2), "mm"),
   sample_color_list=sample_color_list)
hm11_drawn <- ComplexHeatmap::draw(hm11)
```

### Center within batch

```{r hm_centerby_batch_2}
hm12 <- heatmap_se(
   seb,
   column_title="Within-Batch Centering",
   top_colnames=c("Genotype", "Treatment", "batch"),
   column_split=c("Genotype", "Treatment"),
   centerby_colnames="batch",
   row_split=6,
   column_gap=grid::unit(c(2, 4, 2), "mm"),
   sample_color_list=sample_color_list)
hm12_drawn <- ComplexHeatmap::draw(hm12)
```

### Compare Two Heatmaps

* `ComplexHeatmap::Heatmap` objects can be `+` added  
together to display both heatmaps side-by-side.
* **It requires both heatmaps to have identical rows.
   * It does not require the heatmaps to have identical
   clustering, row split.
* `ht_gap` uses `grid::unit()` to define a 3-cm  
gap between heatmaps.

```{r hm_compare_centering}
ComplexHeatmap::draw(
   hm11 + hm12,
   column_title="Comparing Two Data Centering Approaches",
   column_title_gp=grid::gpar(fontsize=20),
   ht_gap=grid::unit(3, "cm")
)
```


# Statistical Hits

## Display Statistical Hits

* `sestats` enables statistical hits on the left side.
* It can be a `list` named by contrast, each with:
   * a `numeric` vector named by `gene` rownames.
   * values should be `-1` down, or `+1` up.
   * Any non-zero value is considered a 'hit'.

* It can be an incidence matrix.
   * `rownames` should match some heatmap `rownames`.
   * `colnames` should be contrast names
   * values are `numeric`, where `-1` is down, `+1` is up.

* It can be `SEStats` from `se_contrast_stats()`.
   * This step is outside the scope, but is convenient!
   * See [Run Limma Statistics](#Run Limma Statistics) for an example.

We create a `list` of hit vectors:

```{r hm_sestats_list_hits}
# define a random set of hits
sestats_list <- list(
   contrast1=setNames(
      sample(c(1, -1), replace=TRUE, size=50),
      sample(rownames(se), size=50)
   ),
   contrast2=setNames(
      sample(c(1, -1), replace=TRUE, size=50),
      sample(rownames(se), size=50)
   )
)
```

* Now create the heatmap with `sestats`:

```{r hm_sestats_list}
hm6 <- heatmap_se(se,
   controlSamples=use_controlSamples,
   control_label="vs groupA",
   sestats=sestats_list,
   column_split="group",
   row_split="Class",
   rowData_colnames="Class",
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm6,
   column_title=attr(hm6, "hm_title"))
```


## Run Limma Statistics

* `se_contrast_stats()` automates `limma` for multiple contrasts.
   * It applies `lmFit()`, `fitContrasts()`, `topTable()`,  
   even `voom()` or `DEqMS` as appropriate.
   * It analyzes each contrast and assay.
   * It returns `SEStats`, used directly in the heatmap

```{r hm_se_contrast_stats_prep}
# SEDesign
sedesign <- groups_to_sedesign(
   colData(se)[, "group", drop=FALSE])
# use only "versus groupA"
contrast_names(sedesign) <- jamba::vigrep("-groupA", contrast_names(sedesign))

# run stats
sestats <- se_contrast_stats(
   se=se,
   fold_cutoff=4,
   sedesign=sedesign,
   assay_name="counts")
```

* Use `SEStats` with `heatmap_se()`:

```{r hm_se_contrast_stats}
hm6s <- heatmap_se(se,
   controlSamples=use_controlSamples,
   control_label="vs groupA",
   sestats=sestats,
   column_split="group",
   row_split=6,
   rowData_colnames="Class",
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm6s,
   column_title=attr(hm6s, "hm_title"))
```

## Drill Down by Row Cluster

* `row_subcluster="4"` makes it easy to answer  
"What is going on with Cluster 4?"

```{r hm_row_subcluster}
# for fun, "drill down" into cluster 5
hm6s_4 <- heatmap_se(se,
   controlSamples=use_controlSamples,
   control_label="vs groupA",
   sestats=sestats,
   column_split="group",
   row_split=6,
   row_subcluster=4,
   rowData_colnames="Class",
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm6s_4,
   column_title=attr(hm6s_4, "hm_title"))
```

## Use a Hit Incidence Matrix

* What is an incidence matrix, you wonder?
   * We demonstrate the data format below.
   * Rows are genes from `rownames(se)`.
   * Columns are contrasts, comparisons, or any useful label.
   * Any non-zero value is considered a hit,  
   and it colorized by direction.

```{r hm_sestats_incidence_hitim, eval=jamba::check_pkg_installed("venndir")}
# convert sestats to list, then incidence matrix
sestats_hitlist <- hit_array_to_list(sestats)
sestats_hitim <- venndir::list2im_value(sestats_hitlist[1:2])
print(head(sestats_hitim));
```

* Now `sestats_hitim` defines statistical hits.

```{r hm_sestats_incidence, eval=jamba::check_pkg_installed("venndir")}
hm7 <- heatmap_se(se,
   controlSamples=use_controlSamples,
   control_label="vs groupA",
   sestats=sestats_hitim,
   column_split="group",
   rowData_colnames="Class",
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm7,
   column_title=attr(hm7, "hm_title"))
```

# Correlation Heatmaps

Any standard `heatmap_se()` can be converted to a  
correlation heatmap by adding:  
`correlation=TRUE`

* It uses the centered data that would have been displayed  
in the heatmap, and instead calls `cor()` to create Pearson  
correlation `matrix`.
   * You can review the data used for correlation using:  
   `hm <- heatmap_se(se)`  
   then:  
   `hm@matrix`
   * Correlation uses `use="pairwise.complete.obs"` and `...`   
   ellipses can be used for custom `method`.

* The heatmap defaults are changed:
   * `max_color=1` so the color range becomes `-1 to +1`
   * `row_split` and `row_label_colname` refer to `colData(se)`.
   * `legend_labels` use `legend_at` directly, no fold conversion.
   * `cluster_columns` controls rows and columns, for symmetry.
   * The default legend title adds `'correlation of'`.


```{r hm_correlation}
hm8corr <- heatmap_se(seb,
   correlation=TRUE,
   apply_hm_column_title=TRUE,
   # controlSamples=use_controlSamples,
   cluster_columns=FALSE,
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm8corr)
```

> Hint: This correlation heatmap shows what we expect  
when there is a sample pairing or batch effect.  
The key feature is "positive correlation along a diagonal."

* You can adjust for pairing factor using  
`centerby_colnames="batch"`:

```{r hm_correlation_batch}
hm8corr_batch <- heatmap_se(seb,
   correlation=TRUE,
   centerby_colnames="batch",
   apply_hm_column_title=TRUE,
   cluster_columns=FALSE,
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm8corr_batch)
```

> Incidentally, the very small negative correlation along  
the diagonal is also indicative of simple batch adjustment.  
The common signal is removed across replicates in the same  
batch, which by definition leaves only slight differences.  
As a result, the residual is typically slightly negative.

> Notice also that the overall group correlations have  
increased substantially, due to the reduction in non-group  
correlated signal. This increase is another indication that  
the batch/pairing effect was appropriately modeled.

# Custom Visual Options


## Add Callout Labels for Some Rows

* `mark_rows` adds labels to the right
   * Great for labeling genes of interest, especially  
   when there are too many rows in the heatmap.
   * We add 5 labels from 'A', and 3 labels from 'D'.

* In RStudio, you may want to silence some warnings:  
`ComplexHeatmap::ht_opt(message=FALSE)`

```{r hm_mark_rows}
# add "callout" labels for a subset of rows
hro <- jamba::heatmap_row_order(hm4);
mark_rows <- c(
   sample(hro[["A"]], size=5),
   sample(hro[["D"]], size=3));

hm5 <- heatmap_se(se,
   mark_rows=mark_rows,
   controlSamples=use_controlSamples,
   control_label="vs groupA",
   column_split="group",
   column_title_rot=90,
   row_split="Class",
   rowData_colnames="Class",
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm5,
   column_title=attr(hm5, "hm_title"))
```


## Customize Column Labels

* Many labels options can be customized: size, color, fontface
   * This example shows how to change the color  
   using `column_names_gp`, which uses `grid::gpar()`.

```{r hm_column_fonts}
# customize column label fonts using column_names_gp
column_bold <- ifelse(
   SummarizedExperiment::colData(se)$group %in% "groupA",
   2, 1);
column_colors <- sample_color_list$group[
   as.character(SummarizedExperiment::colData(se)$group)
]

hm8 <- heatmap_se(se,
   controlSamples=use_controlSamples,
   control_label="vs groupA",
   column_names_gp=grid::gpar(
      col=column_colors,
      font=column_bold
   ),
   column_split="group",
   row_split="Class",
   rowData_colnames="Class",
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm8,
   column_title=attr(hm8, "hm_title"))
```

## Create a Correlation Heatmap

* Just add `correlation=TRUE`.
   * Any heatmap becomes a correlation heatmap,  
   using the same centered data as displayed  
   in the normal heatmap.
   * Centered data are used with `cor()`, which is  
   the recommended way to calculate correlation values.  
   * It answers "What correlates using difference-from-average?"

* `cluster_columns=TRUE` enables column clustering.
* Remove `column_split` to allow unsupervised clustering.


```{r hm_correlation_2}
# correlation=TRUE, any heatmap becomes a sample correlation heatmap
hm8corr <- heatmap_se(se,
   correlation=TRUE,
   apply_hm_column_title=TRUE,
   controlSamples=use_controlSamples,
   control_label="vs groupA",
   column_names_gp=grid::gpar(
      col=column_colors,
      font=column_bold),
   cluster_columns=TRUE,
   sample_color_list=sample_color_list)

ComplexHeatmap::draw(hm8corr)
```



