---
title: "How to Make a Heatmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Make a Heatmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(jamses)
```

# How to Make a Heatmap

> **Still in Progress**
>
> This document is still in draft form, and is being actively
> updated.

The `jamses` package provides `heatmap_se()` which is intended
to "make heatmaps easy" when used with a `SummarizedExperiment`
object.

You can use a `numeric` matrix, but it won't have many of the
fancy extra features.

## About SummarizedExperiment Objects

(See [SummarizedExperiment vignette](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html)
for more details on the object design.)

A `SummarizedExperiment` is just a fancy matrix, with these additions:

* It can store more than one `numeric` matrix.
* It stores `SummarizedExperiment::colData()` with column
annotations.
* It stores `SummarizedExperiment::rowData()` with row
annotations.

Column annotations are used by `heatmap_se()`:

* Column values can split the heatmap.
* Column values can define row-centering subgroups.
* Column annotations can be colorized at the top of a heatmap.

Row annotations are used by `heatmap_se()`:

* Rows can be used to split the heatmap.
* Rows can be used to label specific rows.
* Row annotations can be colorized beside the heatmap.

Several types of data can be converted to `SummarizedExperiment`,
including `ExpressionSet`, `DGEList`, `DESeqDataset`.


## Design Philosophy

The `heatmap_se()` is "opinionated".
Default behavior should make a heatmap very close to ideal.

Most options have exceptions.
There are a lot of exceptions.

We start with `ComplexHeatmap::Heatmap()`. It is amazing.
And complex.
`heatmap_se()` tries to make it easy.


* Data are assumed to be log2-transformed.
* Data are row-centered by default.

   * Data are displayed as row-centered values by default.
   It therefore shows "log2 differences", which are
   equivalent to saying "log2 fold changes".

* **Data are not row-scaled.**

   * Magnitude is an important component of Omics data.
   It is helpful for critical review of the data,
   which is the main purpose of a heatmap.
   * Row-scaling attempts to show patterns relative
   to the variability on each row.
   Ultimately, row-scaling is straightforward, it does not require
   transformation or centering. We feel that it obscures
   important information about the signal:noise in *most*
   Omics datasets.
   For low signal, it assumes that the pattern and noise
   scale together.
   It assumes the patterns are therefore not noise.
   It attempts to allow large and small magnitude effects
   to have similar influence, which is an important goal.
   We aim to achieve this goal without scaling.

* The color scale is blue-white-red, white at zero.

   * Red is `'Up'` because this is a *heat* map.

* The color scale uses a fixed-range.

   * The default heatmap shows a consistent color profile,
   based on biologically relevant changes for *most* Omics
   experiments.
   * The default range spans -8 to +8 fold.
   * Values outside the range appear as the max color.
   * The range may be expanded as needed.
   * The fixed-range is intended to avoid affects by outliers,
   and to avoid over-emphasizing small changes.

* Data centering is a core process:

   * Centering is performed "versus controls".
   * By default, all samples are the controls.
   * Centering is performed within "centerby groups".
   * By default, there is one centerby group.
   * Examples of centerby groups:

      * design factor(s)
      * sample type
      * tissue type
      * cell line
      * processing batch
      * pairing factor

   * Each centerby group may have its own controls.
   For example, you may use "Time 0" as the control
   group for a time series with "Time 1", "Time 2",
   "Time 3", etc.

* Row and Column annotations are "easy".

   * `top_colnames` adds annotations to the top.
   * `rowData_colnames` adds annotations to the left.

* Row `'stat hits'` are "easy" with `sestats`.

   * `sestats` holds statistical hits as an
   as incidence `matrix`, a `list` of rows, or a
   `list` of numeric `sign()` values named by rownames.
   * Hits are shown on the left, colored by direction.
   * By default, the heatmap shows only the hits.

* Rows are grouped using `row_split`.

   * `integer` to split into N groups.
   * `character` to pick annotation columns.

* Columns are grouped using `column_split`.
* Row and column order is "easy".

   * `jamba::heatmap_row_order()`
   * `jamba::heatmap_column_order()`


## Heatmap Basics
